---
title: 'STAT 564: Bayesian Statistics for the Social Sciences'
subtitle: 'Homework 1'
author: "Palmy Chomchat Silarat"
date: "4/3/2022"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
library(formatR)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)

library(rethinking)
```

## 2M1

Compute and plot the grid approximate posterior distribution for each sets of observations.\newline
1. WWW, 2. WWWL, 3. LWWLWWW

```{r}
#I'll try using grid with 1000 points
p_grid <- seq(from=0, to=1, length.out=1000)
prior <- rep(1,1000)
#WWW
likelihood <- dbinom(3, size = 3, prob = p_grid)
posterior_unstd <- prior * likelihood
posterior <- posterior_unstd/sum(posterior_unstd)
#WWWL
likelihood2 <- dbinom(3, size = 4, prob = p_grid)
posterior_unstd2 <- prior*likelihood2
posterior2 <- posterior_unstd2/sum(posterior_unstd2)
#LWWLWWW
likelihood3 <- dbinom(5, size = 7, prob = p_grid)
posterior_unstd3 <- prior*likelihood3
posterior3 <- posterior_unstd3/sum(posterior_unstd3)

par(mfrow=c(1,3))
plot(p_grid, posterior, type = "l", xlab = "Probability of Water", ylab = "Posterior probability", main = "W W W")
plot(p_grid, posterior2, type = "l", xlab = "Probability of Water", ylab = "Posterior probability", main = "W W W L")
plot(p_grid, posterior3, type = "l", xlab = "Probability of Water", ylab = "Posterior probability", main = "L W W L W W W")

```

## 2M2

Assume a prior for p that is equal to zero when p < 0.5 and is a positive constant when p >= 0.5. Compute and plot the grid approximate posterior distribution for each of the sets of observations in the problem just above.

```{r}
#we just need to modify the prior
prior <- ifelse(p_grid < 0.5, 0, 1)
likelihood <- dbinom(3, size = 3, prob = p_grid)
posterior_unstd <- prior * likelihood
posterior <- posterior_unstd/sum(posterior_unstd)
#WWWL
likelihood2 <- dbinom(3, size = 4, prob = p_grid)
posterior_unstd2 <- prior*likelihood2
posterior2 <- posterior_unstd2/sum(posterior_unstd2)
#LWWLWWW
likelihood3 <- dbinom(5, size = 7, prob = p_grid)
posterior_unstd3 <- prior*likelihood3
posterior3 <- posterior_unstd3/sum(posterior_unstd3)

par(mfrow=c(1,3))
plot(p_grid, posterior, type = "l", xlab = "Probability of Water", ylab = "Posterior probability", main = "W W W")
plot(p_grid, posterior2, type = "l", xlab = "Probability of Water", ylab = "Posterior probability", main = "W W W L")
plot(p_grid, posterior3, type = "l", xlab = "Probability of Water", ylab = "Posterior probability", main = "L W W L W W W")


```

## 2M3

Suppose we have 2 globes: Earth, Mars.  \newline
P(W|E) = 0.7, P(W|M) = 0. P(L|E) = 0.3, P(L|M) = 1 \newline
P(E) and P(M) = 0.5 \newline
what is p(E|L)? \newline

Bayes' rules: Posterior = likelihood*Prior/Marginal 

$$
Pr(Earth | Land) = \frac{Pr(Land | Earth) Pr(Earth)}{Pr(Land)}
$$
$$
Pr(Earth | Land) = Pr(Earth) *\frac{Pr(Land | Earth)}{Pr(Land | Earth) + Pr(Land | Mars)}
$$
$$
Pr(Earth | Land) = \frac{0.3 * 0.5}{0.3*0.5+1*0.5} = 0.23
$$
```{r}
#here's the code
P_LE <- 0.3
P_E <- 0.5
P_LM <- 1

posterior <- (0.3*0.5)/(0.3*0.5+1*0.5)
```


## 3E1 - 3E6 

3E1: How much posterior probability lies below p = 0.2? ANSWER = 0.0004 \newline
3E2: How much posterior probability lies below p = 0.8? ANSWER = 0.8884 \newline
3E3: How much posterior probability lies between p = 0.2 and p = 0.8? ANSWER = 0.888 \newline
3E4 20% of the posterior probability lies below which value of p? ANSWER = 0.52 \newline
3E5. 20% of the posterior probability lies above which value of p? ANSWER = 0.76 \newline
3E6. Which values of p contain the narrowest interval equal to 66% of the posterior probability? Answer = 0.51, 0.77 



```{r}
options(scipen=999)
p_grid <- seq( from =0, to=1, length.out = 1000)
prior <- rep(1, 1000)
likelihood <- dbinom(6, size = 9, prob=p_grid)
posterior <- likelihood * prior
posterior <- posterior/sum(posterior)

set.seed(100)
samples <- sample(p_grid, prob = posterior, size=1e4, replace = T)

E1answer <- mean(samples < 0.2)
E2answer <- mean(samples < 0.8)
E3answer <- E2answer - E1answer
E4answer <- quantile(samples, 0.2)
E5answer <- quantile(samples, 0.8)
E6answer <- HPDI(samples, prob = 0.66)



```

## 3M1

Suppose the globe tossing data had turned out to be 8 water in 15 tosses. Construct the posterior distribution, using grid approximation. Use the same flat prior as before.


```{r}
likelihood_3M1 <- dbinom(8, size = 15, prob = p_grid)
post_unstd_3M1 <- likelihood_3M1 * prior
post_3M1 <- post_unstd_3M1/sum(post_unstd_3M1)
plot(p_grid, post_3M1, type = "l", xlab = "Probability of Water", ylab = "Posterior Probability")

```

## 3M2

Draw 10,000 samples from the grid approximation from above. Then use the samples to calculate the 90% HPDI for p.\newline

Answer: The 90th HPDI is between p = 0.34 and 0.73
```{r}
samples <- sample(p_grid, size = 1e4, prob = post_3M1, replace = TRUE)
HPDI(samples, prob = 0.9)

```

## 3M3

Construct a posterior predictive check for this model and data. This means simulate the distribution of samples, averaging over the posterior uncertainty in p. What is the probability of observing 8 water in 15 tosses? \newline

Answer: 0.15 or 15%

```{r}
w_3M3 <- rbinom(1e4, size = 15, prob = samples)
sum(w_3M3 == 8)/1e4

```

## 3M4

Using the posterior distribution constructed from the new (8.15) data, calculate the probability of observing 6 water in 9 tosses. \newline

Answer: 0.18 or 18%

```{r}
w_3M4 <- rbinom(1e4, size = 9, prob = samples)
sum(w_3M4 == 6)/1e4

```

## 3M5 

Start over at 3M1, but now use a prior that is zero below p=0.5 and a constant above p=0.5. This corresponds to prior information that a majority of the Earth’s surface is water. Repeat each problem above and compare the inferences. What difference does the better prior make? If it helps, compare inferences (using both priors) to the true value p=0.7.

```{r}
newprior <- ifelse(p_grid < 0.5, 0, 1) #constant for p above 0.5
likelihood_3M1 <- dbinom(8, size = 15, prob = p_grid)
post_unstd_3M1 <- likelihood_3M1 * newprior
post_3M1 <- post_unstd_3M1/sum(post_unstd_3M1)
plot(p_grid, post_3M1, type = "l", xlab = "Probability of Water", ylab = "Posterior Probability")
```

The 90th HPDI is between p = 0.50 and 0.71 \newline
Code:
```{r}
samples <- sample(p_grid, size = 1e4, prob = post_3M1, replace = TRUE)
HPDI(samples, prob = 0.9)
```

Probability of observing 8 water in 15 tosses = 16% \newline
Code:

```{r}
w_3M3 <- rbinom(1e4, size = 15, prob = samples)
sum(w_3M3 == 8)/1e4
```
Probability of observing 6 water in 9 tosses = 23% \newline
Code:

```{r}
w_3M4 <- rbinom(1e4, size = 9, prob = samples)
sum(w_3M4 == 6)/1e4
```
